The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2014-09-06 Papers/Universalis/ModalCollapse.tex
Make small improvements in the paper

\thanks{This work was completed with the support of ToDo}

@@

\begin{abstract}
The \emph{modal collapse} that afflicts G\"odel's modal ontological 
argument for God's existence is discussed from the perspective of the 
modal square of opposition. Furthermore, a computer-assisted verification 
of the claims that the emendations by Anderson and by Frode are immune to 
the modal collapse is presented.
\end{abstract}

@@

G\"{o}del defines God (see Fig.~\ref{fig1}) as a being who possesses all \emph{positive}
properties.  He does not extensively discuss what positive properties
are, but instead he states a few reasonable (but debatable) axioms
that they should satisfy.  
The overall idea of G{\"o}del's proof is in the tradition of Anselm's
argument, who defined God as some entity of which nothing greater can be
conceived. Anselm argued that existence in the actual world would
make such an assumed being even greater; hence, by definition God must
exist. G{\"o}del's ontological argument is clearly related to this
reasoning pattern. However, it also tries to fix some fundamental
weaknesses in Anselm's work. For example, G{\"o}del explicitly proves
that God's existence is possible, which has been a basic assumption for
Anselm. Because of this, Anselm's argument has been
criticized as incomplete by Leibniz. Leibniz claimed that the
assumption should be derivable from the definition of God as a perfect
being and from the notion of perfection.  G{\"o}del's proof 
addresses this critique, and it also addresses Kant's objection that existence should not be treated as a predicate.

Nevertheless, G{\"o}del's work
still leaves room for criticism. In particular, his axioms are so
strong that they entail a \emph{modal collapse} 
\cite{ToDo:sobel1987?,sobel2004logic}: 
everything that is the case is so necessarily. 
There has been an impressive body of recent and ongoing
work (cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,ToDo:FrodeAndHajek,ContemporaryBibliography} and the references therein)
proposing solutions for the modal collapse. 
The goal of this paper is to present a computer-assisted analysis of 
the solutions proposed by Anderson and by Frode. The technique enabling this analysis is the embedding of quantified modal logics into higher-order logics \cite{J23,B9,C36}, for which automated theorem provers exist \cite{ToDo:leo,sattalax,isabelle,coq}. This technique has already been succesfully employed in the verification and reconstruction of G\"odel's proof \cite{todo:all our previous papers}, and a detailed mathematical description of the technique is available in \cite{todo:ECAI}.

\section{The Modal Collapse}

@@

>>>>

%\thanks{This work was completed with the support of ToDo}

@@

\begin{abstract}
The \emph{modal collapse} that afflicts G\"odel's modal ontological 
argument for God's existence is discussed from the perspective of the 
modal square of opposition.
\end{abstract}

@@

G\"{o}del defines God as a being who possesses all \emph{positive}
properties and states a few reasonable (but debatable) axioms
that such properties should satisfy.  
The overall idea of G{\"o}del's proof is in the tradition of Anselm's
argument, who defined God as some entity of which nothing greater can be
conceived. Anselm argued that existence in the actual world would
make such an assumed being even greater; hence, by definition, God must
exist. However, for Anselm existence was treated as a predicate and the possibility of God's existence was assumed as granted. These issues were criticized by Kant and Leibniz, respectively, and succesfully addressed by G\"odel.

Nevertheless, G{\"o}del's work
still leaves room for criticism. In particular, his axioms are so
strong that they entail a \emph{modal collapse} 
\cite{ToDo:sobel1987?,sobel2004logic}: 
everything that is the case is so necessarily. 
There has been an impressive body of recent and ongoing
work (cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,ToDo:FrodeAndHajek,ContemporaryBibliography} and the references therein)
proposing solutions for the modal collapse. 
The goal of this short note is merely to remark that the modal collapse is a collapse of the modal square of opposition.

\section{The Collapse of the Modal Square}

@@

The technique enabling this analysis is the embedding of quantified modal logics into higher-order logics \cite{J23,B9,C36}, for which automated theorem provers exist \cite{ToDo:leo,sattalax,isabelle,coq}. This technique has already been succesfully employed in the verification and reconstruction of G\"odel's proof \cite{todo:all our previous papers}, and a detailed mathematical description of the technique is available in \cite{todo:ECAI}.
